Measuring causal influence with\\ back-to-back regression: the linear case

### 2019-09-24 paper/synthetic.tex
no changes, just line length set to 80

The data generating process for each experiment constructs $n$ training examples according to the model $Y = (\text{snr} \cdot XE + N)F$, where $\text{snr}$ is a scalar that modulates the signal-to-noise ratio.
%
Here,
%\begin{itemize}
%    \item $F \in \mathbb{R}^{d_x \times d_y}$ contains entries drawn from $\mathcal{N}(0, d_x^{-1})$,
%    \item $X \in \mathbb{R}^{n \times d_x}$ contains rows drawn from $\mathcal{N}(0, \Sigma_X)$,
%    \item $N \in \mathbb{R}^{n \times d_x}$ contains rows drawn from $\mathcal{N}(0, \Sigma_N)$,
%    \item $E \in \mathbb{R}^{d_x \times d_x}$ is a binary diagonal matrix containing $n_c$ ones,
%    \item $\Sigma_X = AA^\top$, where $A \in \mathbb{R}^{d_x \times d_x}$ contains entries drawn from $\mathcal{N}(0, d_x^{-1})$,
%    \item $\Sigma_N = BB^\top$, where $B \in \mathbb{R}^{d_x \times d_x}$ contains entries drawn from $\mathcal{N}(0, d_x^{-1})$,
%\end{itemize}
    $F \in \mathbb{R}^{d_x \times d_y}$ contains entries drawn from $\mathcal{N}(0, d_x^{-1})$,
    $X \in \mathbb{R}^{n \times d_x}$ contains rows drawn from $\mathcal{N}(0, \Sigma_X)$,
    $N \in \mathbb{R}^{n \times d_x}$ contains rows drawn from $\mathcal{N}(0, \Sigma_N)$,
    $E \in \mathbb{R}^{d_x \times d_x}$ is a binary diagonal matrix containing $n_c$ ones,
    $\Sigma_X = AA^\top$ where $A \in \mathbb{R}^{d_x \times d_x}$ contains entries drawn from $\mathcal{N}(0, d_x^{-1})$,
    $\Sigma_N = BB^\top$ where $B \in \mathbb{R}^{d_x \times d_x}$ contains entries drawn from $\mathcal{N}(0, d_x^{-1})$,
and the factor $\text{snr} \in (0, \infty)$.

@@

We compare the performance of B2B against five competing methods:
%
Ridge regression \citep{hoerl1959optimum}, multitask Lasso \citep{argyriou2008convex}, Partial Least Squares or PLS \citep{wold_pls, tenenhaus_pls}, Canonical Correlation Analysis or CCA \citep{cca_hotelling}, and Reduced Rank Regression or RRR \citep{Izenman_rrr}.
%
Each of these methods estimates a matrix of coefficients $\hat{W} \in \mathbb{R}^{d_x \times d_y}$, from which we estimate the diagonal mask $\hat{E}$ using the feature importances $\hat{E}_{i,i} = \| W_{i, :} \|$.
%
For B2B, we obtain $\hat{E}$ as described in Algorithm~\ref{algorithm:b2br}.
%
We employ $5$-fold cross-validation to select the optimal hyper-parameters of each method.

>>>>

The data generating process for each experiment constructs $n$ training examples
according to the model $Y = (\text{snr} \cdot XE + N)F$, where $\text{snr}$ is a
scalar that modulates the signal-to-noise ratio.
%
Here,
%\begin{itemize} \item $F \in \mathbb{R}^{d_x \times d_y}$ contains entries
%drawn from $\mathcal{N}(0, d_x^{-1})$, \item $X \in \mathbb{R}^{n \times d_x}$
%contains rows drawn from $\mathcal{N}(0, \Sigma_X)$, \item $N \in \mathbb{R}^{n
%\times d_x}$ contains rows drawn from $\mathcal{N}(0, \Sigma_N)$, \item $E \in
%\mathbb{R}^{d_x \times d_x}$ is a binary diagonal matrix containing $n_c$ ones,
%\item $\Sigma_X = AA^\top$, where $A \in \mathbb{R}^{d_x \times d_x}$ contains
%entries drawn from $\mathcal{N}(0, d_x^{-1})$, \item $\Sigma_N = BB^\top$,
%where $B \in \mathbb{R}^{d_x \times d_x}$ contains entries drawn from
%$\mathcal{N}(0, d_x^{-1})$, \end{itemize}
    $F \in \mathbb{R}^{d_x \times d_y}$ contains entries drawn from
$\mathcal{N}(0, d_x^{-1})$, $X \in \mathbb{R}^{n \times d_x}$ contains rows
drawn from $\mathcal{N}(0, \Sigma_X)$, $N \in \mathbb{R}^{n \times d_x}$
contains rows drawn from $\mathcal{N}(0, \Sigma_N)$, $E \in \mathbb{R}^{d_x
\times d_x}$ is a binary diagonal matrix containing $n_c$ ones, $\Sigma_X =
AA^\top$ where $A \in \mathbb{R}^{d_x \times d_x}$ contains entries drawn from
$\mathcal{N}(0, d_x^{-1})$, $\Sigma_N = BB^\top$ where $B \in \mathbb{R}^{d_x
\times d_x}$ contains entries drawn from $\mathcal{N}(0, d_x^{-1})$, and the
factor $\text{snr} \in (0, \infty)$.

@@

We compare the performance of B2B against four competing methods, all
implemented in scikit-learn \citep{scikit}:
%
% To be updated

\subsection{Baseline models}

Forward regression consists of an $l2$-regularized "ridge" regression from the
putative causes $X$ to the observations $Y$: \begin{equation} H = (X^T X
+\lambda I)^{-1} X^T Y \end{equation}

Backward regression consists of an $l2$-regularized "ridge" regression from $Y$
to $X$: \begin{equation} G = (Y^T Y +\lambda I)^{-1} Y^T X \end{equation}

CCA finds $G\in\mathbb{R}^{d_z, d_y}$ and $H\in\mathbb{R}^{d_z, d_x}$ such that
$X$ and $Y$ are maximally correlated in a latent $Z$ space: \begin{equation} max
corr(XH^T, YG^T) \end{equation}

% To be checked
PLS finds $G\in\mathbb{R}^{d_z, d_y}$ and $H\in\mathbb{R}^{d_z, d_x}$ such that
$X$ and $Y$ are maximally covarying in a latent $Z$ space: \begin{equation} max
cov(XH^T, YG^T) \end{equation}

We employ $5$-fold cross-validation to select the optimal number of components
for CCA and PLS. Regressions were l2-regularized with a $\lambda$ regularization
parameters fitted with the efficient leave-one-out procedure implemented in
scikit-learn RidgeCV \citep{scikit}.

\subsection{Evaluating Causal Discovery from models' coefficients}

B2B leads to unbiased (i.e. zeros-centered) scalar coefficients for non-causal
features. In contrast, the Forward, Backward, CCA and PLS models lead to a
loading vector $H_i$ per feature $i$ (or one vector $G^i$ for the backward
model). To transform such vector into an estimated causal contribution $\hat E$,
we take the sum of square coefficients:

\begin{equation} \hat E_i = \sum_j {H^j_i}^2 \end{equation}

To estimate whether models accurately identify causal factors, we compute the
area-under-the-curve (AUC) across factors $AUC(E, \hat E)$.
%\begin{equation} AUC = 1 - \sum_1^n (E_k - E_{k-1}) ( \hat{E}_k +
%\hat{E}_{k-1}) / 2 \end{equation}

\subsection{Evaluating Causal Discovery with held-out prediction reliability}

In most cases, $E$ is not known and AUC can thus not be estimated.

To address this issue, we assess the ability of each model to reliably predict
independent and identically distributed data from $Y$, given all of the $X$
features versus all but one feature: $X_k$ (i.e. 'knock-out X'). This procedure
results in two correlation metrics $R_{full}$ and $R_{knockout}$, whose
difference $\Delta R=R_{full}-R_{knockout}$ indicates how much $X_i$ improves
the prediction of $Y$ above and beyond what can be predicted from all other
features $X_k$.

For, the Forward Model, this feature importance is assessed as follow:

\begin{algorithm}[H]
    %\SetAlgoLined
    \KwIn{$X_{train} \in \mathbb{R}^{n \times d_x}$, $X_{test} \in
\mathbb{R}^{n' \times d_x}$, $Y_{train} \in \mathbb{R}^{n\times d_y}$, $Y_{test}
\in \mathbb{R}^{n'\times d_y}$, } \KwOut{estimate of prediction improvement
$\Delta{R} \in \mathbb{D}^{d_x, d_y}$.} $H = \text{LinearRegression}(X_{train},
Y_{train})$\; $R_{full} = \text{corr}(X_{test} H, Y_{test})$\;

    \For{$i = 1, \ldots, d_x$}{ $K = Id$\; $K[i] \leftarrow 0$\; $H_k =
\text{LinearRegression}(X_{train} K, Y_{train})$\; $R_{k} = \text{corr}(X_{test}
{H_k}, Y_{test})$\; $\Delta R_i = R_{full} - R_{k}$\; } \Return{$\Delta R$}
\caption{Forward feature importance.} \label{algorithm:fwd_fi} \end{algorithm}

\iffalse \begin{enumerate} \item Fit $H$ given $X_{train}$ and $Y_{train}$ \item
For each feature $i$: \begin{itemize} \item Define $K$, and identity matrix
whose row $i$ has been zeroed-out \item Fit $H^i_k$ given $X_{train} K$ and
$Y_{train}$ \end{itemize} \end{enumerate} \fi

\begin{algorithm}[H]
    %\SetAlgoLined
    \KwIn{$X_{train} \in \mathbb{R}^{n \times d_x}$, $X_{test} \in
\mathbb{R}^{n' \times d_x}$, $Y_{train} \in \mathbb{R}^{n\times d_y}$, $Y_{test}
\in \mathbb{R}^{n'\times d_y}$, } \KwOut{estimate of prediction improvement
$\Delta{R} \in \mathbb{D}^{d_x, d_z}$.} $H, G = \text{CCA}(X_{train},
Y_{train})$\; $R_{full} = \text{corr}(X_{test} H, Y_{test} G)$\;

    \For{$i = 1, \ldots, d_x$}{ $K = Id$\; $K[i] \leftarrow 0$\; $H_k, G_k =
\text{CCA}(X_{train} K, Y_{train})$\; $R_{k} = \text{corr}(X_{test} H_k,
Y_{test} G_k)$\; $\Delta R_i = R_{full} - R_{k}$\; } \Return{$\Delta R$}
\caption{CCA and PLS feature importance.} \label{algorithm:cdp_fi}
\end{algorithm}

\iffalse For CCA and PLS, feature importance is assessed as follow:
\begin{enumerate} \item Fit $H$ and $G$ given $X_{train}$ and $Y_{train}$ \item
For each feature $i$: \begin{itemize} \item Define $K$, and identity matrix
whose row $i$ has been zeroed-out \item Fit $H^i_k$ and $G_k$ given $X_{train}
K$ and $Y_{train}$ \end{itemize} \end{enumerate} \fi

For the Backward Model, feature importance cannot be assessed. We thus simply
report the model performance $R_i = corr(YG_i, X_i)$

For, B2B feature importance is assessed as follow:

\begin{algorithm}[H]
    %\SetAlgoLined
    \KwIn{$X_{train} \in \mathbb{R}^{n \times d_x}$, $X_{test} \in
\mathbb{R}^{n' \times d_x}$, $Y_{train} \in \mathbb{R}^{n\times d_y}$, $Y_{test}
\in \mathbb{R}^{n'\times d_y}$, } \KwOut{estimate of prediction improvement
$\Delta{R} \in \mathbb{D}^{d_x}$.} $H, G = \text{B2B}(X_{train}, Y_{train})$\;
$R_{full} = \text{corr}(X_{test} H, Y_{test} G)$\;

    \For{$i = 1, \ldots, d_x$}{ $K = Id$\; $K[i] \leftarrow 0$\; $H_k =
\text{LinearRegression}(X_{train} K, Y_{train} G_i)$\; $R_{k} =
\text{corr}(X_{test} H_k, Y_{test} G_i)$\; $\Delta R_i = R_{full} - R_{k}$\; }
\Return{$\Delta R$} \caption{B2B feature importance.} \label{algorithm:cdp_fi}
\end{algorithm}

\iffalse \begin{enumerate}

  \item Fit $G$ given $X_{train}$ and $Y_{train}$ \item Fit $H$ given
$X_{train}$ and $Y_{train} G^T$ \item For each feature $i$: \begin{itemize}
\item Define $K$, and identity matrix whose row $i$ has been zeroed-out \item
Fit $H^i_k$ given $X_{train} K$ and $Y_{train} G^T$ \end{itemize}

  \item Compute $R_{full}$ given $Y_{test} G^T$ and $X_{test}$ \item For all $i
\in d_x$: \begin{itemize} \item Compute $R^i_{knockout}$ given $Y_{test} G^T$
and $X_{test} K {H^i_k}^T$ \item Compute $R^i_{delta}=R_{full} - R^i_{knockout}$
\end{itemize} \end{enumerate} \fi

\iffalse Ridge regression \citep{hoerl1959optimum}, multitask Lasso
\citep{argyriou2008convex}, Partial Least Squares or PLS \citep{wold_pls,
tenenhaus_pls}, Canonical Correlation Analysis or CCA \citep{cca_hotelling}, and
Reduced Rank Regression or RRR \citep{Izenman_rrr}.
%
Each of these methods estimates a matrix of coefficients $\hat{W} \in
\mathbb{R}^{d_x \times d_y}$, from which we estimate the diagonal mask $\hat{E}$
using the feature importances $\hat{E}_{i,i} = \| W_{i, :} \|$.
%
For B2B, we obtain $\hat{E}$ as described in Algorithm~\ref{algorithm:b2br}.
%
